---
title: "BSE_QuartoFINAL155"
author: "Batsambuu Batbold, Sophia Stevenson, Evelyn Paul"
format:
  html:
    toc: true
    embed-resources: true
    code-tools: true    
---

This code is probably easier to view here: https://github.com/sophiaaasmile/stat-155-project/tree/main

In separate files it's going to be easier to parse this mess of code. 

Also, you need to download this file and access the path from your local desktop https://github.com/sophiaaasmile/stat-155-project/blob/baf9953ae78c20d65d19d1ef6b4f21dbc689033b/US_County_Boundaries.zip
Github doesn't let us upload the unzipped files without crashing the entire repository. When you get to the chunk where we use it we will warn you again, or you will run into an error. 


```{r}
library(ggplot2)
library(broom)
library(readr)
library(tidyr)
library(wesanderson)
library(tibble)
library(arrow)    
library(dplyr)
library(data.table)
library(stringr) 
library(fs) 
library(sf)
library(purrr)
library(leaflet)
library(maps)
library(tigris)
library(rnaturalearth)
library(terra) 
library(tidyterra)
library(lubridate)
library(gridExtra)
library(scales)

knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE, 
                      fig.align = "center", fig.width = 8, fig.height = 6, include = FALSE)
```

#### Crop and GDP ####

```{r}
# Reading the crop and GDP data files
GDP2022 <- read.csv("Data/GDP2022.csv")

wheat2022 <- read.csv("Data/wheat_area_sale_production.csv")
soybean2022 <- read.csv("Data/soybeans_area_sale_production.csv")
corn2022 <- read.csv("Data/corn_area_sales_production.csv")
```

```{r}
# Cleaning and reformatting the GDP dataset: 
#   separate state and county, remove symbols, and match full state names

GDP2022 %>%
  separate(GeoName, into = c("County", "StateAbb"), sep = ",\\s*") %>%
  mutate(StateAbb = gsub("\\*", "", StateAbb)) %>%
  mutate(State = state.name[match(StateAbb, state.abb)]) %>%
  select(-StateAbb, -GeoFips) -> GDP2022
```

```{r}
# Defining a general function to clean crop data: 
#   filter for 2022, reshape, rename columns, and merge with GDP
clean_data <- function(crop_data, crop_name1, crop_name2 = crop_name1, crop_name3 = crop_name1){
  crop_data %>%
    filter(Year == 2022) %>%
    select(State, County, Data.Item, Value) %>%
    mutate(Value = as.numeric(gsub(",", "", Value))) -> crop_clean
  
  crop_clean %>%
    pivot_wider(names_from = Data.Item, values_from = Value) %>%
    rename(area = paste(crop_name1, '- ACRES HARVESTED'), production = paste(crop_name2, '- PRODUCTION, MEASURED IN BU'), sales = paste(crop_name3, '- SALES, MEASURED IN $')) -> crop_clean
  
  crop_clean %>%
    mutate(State = tolower(State), County = tolower(County)) %>%
    full_join(
      GDP2022 %>% mutate(State = tolower(State), County = tolower(County)), 
      by = c("State", "County")
    ) %>% drop_na() %>% mutate(GDP = as.numeric(GDP)) -> crop_clean
  
  return (crop_clean)
}
```

```{r}
# Cleaning individual crop datasets and combine them into a single dataset for analysis
soybean_clean <- clean_data(soybean2022, "SOYBEANS") %>% mutate(crop = "SOYBEANS")
wheat_clean <- clean_data(wheat2022, "WHEAT") %>% mutate(crop = "WHEAT")
corn_clean <- clean_data(corn2022, "CORN, GRAIN", crop_name3 = "CORN") %>% mutate(crop = "CORN")

rbind(corn_clean, soybean_clean, wheat_clean) %>% arrange(State, County) -> data_clean
```

```{r}
# Creating a scatter plot of Sales vs. Production, colored by crop type
data_clean %>%
  ggplot(aes(x = sales, y = production, color = crop)) + 
  geom_point(alpha = 0.6, size = 2) +
  geom_smooth(method = "lm", se = FALSE, linewidth = 1.2) +
  scale_x_continuous(labels = dollar_format()) +
  scale_y_continuous(labels = comma_format()) +
  labs(
    title = "Sales vs. Production by Crop",
    subtitle = "Each point represents a U.S. county in 2022",
    x = "Sales (USD)",
    y = "Production (Bushels)",
    color = "Crop"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    plot.title = element_text(face = "bold", size = 16),
    plot.subtitle = element_text(size = 12, margin = margin(b = 10)),
    axis.title = element_text(face = "bold"),
    axis.text = element_text(color = "gray20"),
    legend.title = element_text(face = "bold"),
    plot.caption = element_text(size = 9, face = "italic", color = "gray40"),
    panel.grid.minor = element_blank()
  )
```

```{r}
# Creating a scatter plot of Area vs. Production, colored by crop type
data_clean %>%
  ggplot(aes(x = area, y = production, color = crop)) + 
  geom_point(alpha = 0.6, size = 2) +
  geom_smooth(method = "lm", se = FALSE, linewidth = 1.2) +
  scale_x_continuous(labels = comma_format()) +
  scale_y_continuous(labels = comma_format()) +
  labs(
    title = "Area vs. Production by Crop",
    subtitle = "Each point represents a U.S. county in 2022",
    x = "Area Harvested (Acres)",
    y = "Production (Bushels)",
    color = "Crop"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    plot.title = element_text(face = "bold", size = 16),
    plot.subtitle = element_text(size = 12, margin = margin(b = 10)),
    axis.title = element_text(face = "bold"),
    axis.text = element_text(color = "gray20"),
    legend.title = element_text(face = "bold"),
    panel.grid.minor = element_blank()
  )
```



```{r}
# Fitting a multiple linear regression model
mod <- lm(GDP ~ production + area + sales + crop, data = data_clean)

summary(mod)
```
```{r}
# Building and saving a table of regression coefficients as an image
coefs <- data.frame(
  Predictor = c("(Intercept)", "Production", "Area", "Sales", "Crop: Soybeans", "Crop: Wheat"),
  Estimate = c(6313000, -0.236, -43.37, 0.04921, -869000, -335500),
  Std_Error = c(397000, 0.1313, 9.527, 0.02439, 600200, 561800),
  t_value = c(15.90, -1.796, -4.553, 2.018, -1.448, -0.597),
  p_value = c("< 0.001", "0.0726", "< 0.001", "0.0437", "0.1477", "0.5504")
)

coefs$Estimate <- prettyNum(round(coefs$Estimate, 3), big.mark = ",")
coefs$Std_Error <- prettyNum(round(coefs$Std_Error, 3), big.mark = ",")
coefs$t_value <- round(coefs$t_value, 2)

tbl <- tableGrob(coefs)
ggsave("regression_table.png", tbl, width = 7, height = 3, dpi = 300)
```


#### NOAA Models for Monthly ####


```{r}
weather <- read_csv("modified_noaa_monthly.csv")

ggplot(weather, aes(x = TDIF, y = PRCP, color = YEAR)) +
  geom_point(alpha=0.05) +
  geom_smooth(method="lm")

ggplot(filter(weather, MONTH == 7), aes(x = ELEV, y = TMAX)) +
  geom_point(alpha=0.05) +
  geom_smooth()

WM0 <- LM(TMAX ~ YEAR * LAT, data = filter(weather, MONTH == 7))
WM1 <- lm(PRCP ~ YEAR * LAT + YEAR * LONG, data = filter(weather, TDIF >= -100))
WM2 <- lm(PRCP ~ TDIF + YEAR * LAT + YEAR * LONG, data = weather)
summary(WM0)
summary(WM1)
summary(WM2)

anova(with(weather, WM1), with(weather,WM2))

summary(lm(TDIF ~ LONG + LAT, data = weather))

```

### Weather combined with crops ###

```{r}

stations <- readRDS(
  url(
    "https://raw.githubusercontent.com/sophiaaasmile/stat-155-project/main/station_list.rds",
    "rb"
  )
)

cotton    <- read_csv("https://raw.githubusercontent.com/sophiaaasmile/stat-155-project/ea18c576a116251cda738ac8bb912a70d86d2ca9/agricsvs/CottonArea_20Yr.csv")
wheat     <- read_csv("https://raw.githubusercontent.com/sophiaaasmile/stat-155-project/ea18c576a116251cda738ac8bb912a70d86d2ca9/agricsvs/WheatArea_20Yr.csv")
corn      <- read_csv("https://raw.githubusercontent.com/sophiaaasmile/stat-155-project/ea18c576a116251cda738ac8bb912a70d86d2ca9/agricsvs/cornArea_20yrcounty.csv")
cornCash  <- read_csv("https://raw.githubusercontent.com/sophiaaasmile/stat-155-project/ea18c576a116251cda738ac8bb912a70d86d2ca9/agricsvs/cornCashYield_20yrcounty.csv")
fruitnuts <- read_csv("https://raw.githubusercontent.com/sophiaaasmile/stat-155-project/ea18c576a116251cda738ac8bb912a70d86d2ca9/agricsvs/fruitTreeNutArea_20yrcounty.csv")
hay       <- read_csv("https://raw.githubusercontent.com/sophiaaasmile/stat-155-project/ea18c576a116251cda738ac8bb912a70d86d2ca9/agricsvs/hayArea_20yrcounty.csv")
soybean   <- read_csv("https://raw.githubusercontent.com/sophiaaasmile/stat-155-project/ea18c576a116251cda738ac8bb912a70d86d2ca9/agricsvs/soybeanArea_20yrcounty.csv")
```

```{r}

weather_cols <- c(                      
  "STATION", "DATE", "LATITUDE", "LONGITUDE", "ELEVATION",
  "TAVG", "TMAX", "TMIN",                # temperature core
  "PRCP",                                # daily precip depth
  "SNDP",                                # snow depth on ground
  "CLDD", "HTDD",                        # cooling / heating degree-days
  "DP01", "DP10", "DP1X",                # days ≥ 0.01 / ≥ 0.10 / ≥ 1 in
  "DX32", "DX70", "DX90",                # days ≥ 32 °F, 70 °F, 90 °F
  "DT32", "DT00",                        # days < 32 °F, < 0 °F
  "EMXT", "EMNT",                        # monthly record high / low
  "EMXP",                                # monthly record precip
  "FRSHTT"                               # fog-rain-snow-hail-thunder-tornado flags
)

all_stations <- stations %>%
  lapply(function(x) {
    x %>%
      mutate(across(everything(), as.character)) %>%
      select(any_of(weather_cols))
  }) %>%
  bind_rows(.id = "STATION") %>%      # ← this turns the list into a tibble
  mutate(
    network = case_when(
      str_starts(STATION, "US1")   ~ "COOP",
      str_starts(STATION, "GHCND") ~ "GHCN",
      str_detect(STATION, "K[A-Z]{3}") ~ "ASOS/AWOS",
      TRUE                          ~ "Other"
    )
  )
```

```{r}
#Clean the data 
#number data should actually be stay numbers. putting it back
num_cols <- c("LATITUDE", "LONGITUDE", "ELEVATION", "TAVG", "TMAX", "TMIN", "PRCP")
all_stations <- all_stations %>%
  mutate(across(any_of(num_cols), ~ {
    nums <- suppressWarnings(as.numeric(.))
    ifelse(is.finite(nums), nums, NA)  
  }))

# Filter like a blue whale 
stations_geo <- all_stations %>% 
  filter(
    !is.na(LATITUDE), 
    !is.na(LONGITUDE),
    between(LONGITUDE, -180, -50),  
    between(LATITUDE, 15, 72)       
  ) %>%
  arrange(STATION, desc(DATE)) %>%
  distinct(STATION, .keep_all = TRUE)

stations_sf <- stations_geo %>%
  st_as_sf(
    coords = c("LONGITUDE", "LATITUDE"),
    crs = 4326,
    remove = FALSE
  ) %>%
  st_make_valid()

```

> !!! UNZIP THIS FILE NOW !!! 
> https://github.com/sophiaaasmile/stat-155-project/blob/baf9953ae78c20d65d19d1ef6b4f21dbc689033b/US_County_Boundaries.zip

```{r}

us_counties <- read_csv("~/Desktop/US_County_Boundaries/US_County_Boundaries.csv") #or whatever your actual path is 
us_counties_sf <- st_read("~/Desktop/US_County_Boundaries/US_County_Boundaries.shp")

```

```{r}

# More cleaning, joins, etc 

us_counties_sf <- st_transform(us_counties_sf, st_crs(stations_sf))

stations_with_county <- st_join(
  stations_sf,
  us_counties_sf %>% select(GEOID),
  join = st_within
)

station2county <- stations_with_county %>%
  st_drop_geometry() %>%
  select(STATION, GEOID)

all_stations <- all_stations %>%
  mutate(
    DATE = as.character(DATE),
    year = as.integer(substr(DATE, 1, 4))
  )

station2county <- stations_with_county %>%
  st_drop_geometry() %>%
  select(STATION, GEOID)

weather_all <- all_stations %>%
  mutate(year = as.integer(DATE)) %>%
  left_join(station2county, by = "STATION") %>%
  filter(!is.na(GEOID))

weather_daily <- all_stations %>%
  mutate(
    year = as.integer(DATE)
  ) %>%
  left_join(station2county, by = "STATION") %>%
  filter(!is.na(GEOID))

weather_daily <- all_stations %>%
  mutate(
    year = as.integer(DATE)
  ) %>%
  left_join(station2county, by = "STATION") %>%
  filter(!is.na(GEOID)) %>%
  mutate(across(
    c(TAVG, TMAX, TMIN, PRCP,
      DP01, DP10, DP1X,
      DX32, DX70, DX90,
      DT32, DT00,
      CLDD, HTDD),
    ~ as.numeric(.)
  ))

weather_summary <- weather_daily %>%
  group_by(GEOID, year) %>%
  summarise(
    mean_tavg       = mean(TAVG, na.rm = TRUE),
    mean_tmax       = mean(TMAX, na.rm = TRUE),
    mean_tmin       = mean(TMIN, na.rm = TRUE),
    total_prcp      = sum(PRCP, na.rm = TRUE),
    prcp_days       = sum(DP01, na.rm = TRUE),
    prcp10_days     = sum(DP10, na.rm = TRUE),
    prcp1in_days    = sum(DP1X, na.rm = TRUE),
    dx32_days       = sum(DX32, na.rm = TRUE),
    dx70_days       = sum(DX70, na.rm = TRUE),
    hot_days        = sum(DX90, na.rm = TRUE),
    freeze32_days   = sum(DT32, na.rm = TRUE),
    freeze0_days    = sum(DT00, na.rm = TRUE),
    cdd_sum         = sum(CLDD, na.rm = TRUE),
    hdd_sum         = sum(HTDD, na.rm = TRUE),
    gdd_base50      = sum(pmax(((TMAX + TMIN)/2) - 50, 0), na.rm = TRUE),
    record_max_temp = max(EMXT, na.rm = TRUE),
    record_min_temp = min(EMNT, na.rm = TRUE),
    record_prcp     = max(EMXP, na.rm = TRUE),
    .groups         = "drop"
  )
```

```{r}

wheat2 <- wheat %>%
  transmute(
    # zero-pad state and county ANSI codes. Credit to geo nerds who figured this out. not me. 
    state_fips  = str_pad(state_ansi, 2, pad = "0"),
    county_fips = str_pad(county_ansi, 3, pad = "0"),
    GEOID       = paste0(state_fips, county_fips),   # e.g. "01031"
    year        = as.integer(year),
    wheat_area  = as.numeric(Value)
  )

cotton2 <- cotton %>%
  transmute(
    state_fips    = str_pad(state_ansi,  2, pad = "0"),
    county_fips   = str_pad(county_ansi, 3, pad = "0"),
    GEOID         = paste0(state_fips, county_fips),
    year          = as.integer(year),
    cotton_area   = as.numeric(Value)
  )

corn2 <- corn %>%
  transmute(
    state_fips   = str_pad(state_ansi,  2, pad = "0"),
    county_fips  = str_pad(county_ansi, 3, pad = "0"),
    GEOID        = paste0(state_fips, county_fips),
    year         = as.integer(year),
    corn_area    = as.numeric(Value)
  )

cornCash2 <- cornCash %>%
  transmute(
    state_fips      = str_pad(state_ansi,  2, pad = "0"),
    county_fips     = str_pad(county_ansi, 3, pad = "0"),
    GEOID           = paste0(state_fips, county_fips),
    year            = as.integer(year),
    corn_cash_yield = as.numeric(Value)
  )

fruitnuts2 <- fruitnuts %>%
  transmute(
    state_fips      = str_pad(state_ansi,  2, pad = "0"),
    county_fips     = str_pad(county_ansi, 3, pad = "0"),
    GEOID           = paste0(state_fips, county_fips),
    year            = as.integer(year),
    fruitnut_area   = as.numeric(Value)
  )

hay2 <- hay %>%
  transmute(
    state_fips   = str_pad(state_ansi,  2, pad = "0"),
    county_fips  = str_pad(county_ansi, 3, pad = "0"),
    GEOID        = paste0(state_fips, county_fips),
    year         = as.integer(year),
    hay_area     = as.numeric(Value)
  )

soybean2 <- soybean %>%
  transmute(
    state_fips    = str_pad(state_ansi,  2, pad = "0"),
    county_fips   = str_pad(county_ansi, 3, pad = "0"),
    GEOID         = paste0(state_fips, county_fips),
    year          = as.integer(year),
    soybean_area  = as.numeric(Value)
  )

```

```{r}
df_mod_wheat <- inner_join(
  wheat2,
  weather_summary,
  by = c("GEOID", "year")
)

df_mod_corn <- inner_join(
  corn2,
  weather_summary,
  by = c("GEOID", "year")
)

df_mod_cornCash <- inner_join(
  cornCash2,
  weather_summary,
  by = c("GEOID", "year")
)

df_mod_cotton <- inner_join(
  cotton2,
  weather_summary,
  by = c("GEOID", "year")
)

df_mod_fruitnuts <- inner_join(
  fruitnuts2,
  weather_summary,
  by = c("GEOID", "year")
)

df_mod_hay <- inner_join(
  hay2,
  weather_summary,
  by = c("GEOID", "year")
)

df_mod_soybean <- inner_join(
  soybean2,
  weather_summary,
  by = c("GEOID", "year")
)


```

### Corn - You can make any of the major crop models but wheat was the only one that we could pull $/Acres in a timely manner (see usdaapiscript) so this our best crop/weather model).L= Log models are slightly better fits than anything else. See cropviz.qmd for the full set of 

```{r}

# Full linear
corn_full_mod <- lm(
  corn_area ~ 
    total_prcp + mean_tavg + hot_days +
    prcp_days + prcp1in_days + freeze32_days +
    cdd_sum + hdd_sum + gdd_base50,
  data      = df_mod_corn,
  na.action = na.exclude
)
summary(corn_full_mod)

# Log‐linear + interaction
df_log_corn <- df_mod_corn %>%
  mutate(
    log_corn  = log(corn_area + 1),
    log_prcp  = log(total_prcp + 1),
    log_temp  = log(mean_tavg   + 1)
  ) %>%
  drop_na()
corn_log_mod <- lm(
  log_corn ~ log_prcp * log_temp +
               hot_days + prcp_days + prcp1in_days +
               freeze32_days + cdd_sum + hdd_sum,
  data = df_log_corn
)
summary(corn_log_mod)

# 4) Residuals vs fitted
resid_df_corn <- augment(corn_log_mod)
ggplot(resid_df_corn, aes(.fitted, .resid)) +
  geom_point(alpha = 0.3) + geom_hline(yintercept = 0) +
  labs(title="Corn Residuals vs Fitted",
       x="Fitted", y="Residuals") +
  theme_minimal()
```

```{r}
# Merge in cash yield & weather
df_mod_corn_cash <- corn2 %>%
  inner_join(cornCash2, by = c("GEOID","year")) %>%
  inner_join(weather_summary, by = c("GEOID","year"))

# Cash + weather model
corn_cash_area_mod <- lm(
  corn_area ~ corn_cash_yield +
    total_prcp + mean_tavg +
    hot_days + prcp_days + prcp1in_days +
    freeze32_days + cdd_sum + hdd_sum,
  data      = df_mod_corn_cash,
  na.action = na.exclude
)
summary(corn_cash_area_mod)

# Log‐linear with cash
df_log_corn_cash <- df_mod_corn_cash %>%
  mutate(
    log_area = log(corn_area       + 1),
    log_cash = log(corn_cash_yield + 1),
    log_prcp = log(total_prcp      + 1),
    log_temp = log(mean_tavg       + 1)
  ) %>%
  drop_na()
corn_cash_log_mod <- lm(
  log_area ~ log_cash + log_prcp * log_temp +
               hot_days + prcp_days + prcp1in_days +
               freeze32_days + cdd_sum + hdd_sum,
  data = df_log_corn_cash
)
summary(corn_cash_log_mod)

```

```{r}

# 1. Set up the BottleRocket2 palette
br2 <- wes_palette("BottleRocket2")

# 2. Scatterplot with linear fit
ggplot(df_mod_corn_cash, aes(x = corn_cash_yield, y = corn_area)) +
  geom_point(alpha = 0.6, color = br2[2]) +
  geom_smooth(method = "lm", se = FALSE, color = br2[4]) +
  labs(
    title = "County-Year Corn Area vs. Cash Yield",
    x     = "Cash Yield ($ per acre)",
    y     = "Corn Area (acres)"
  ) +
  theme_minimal()
```
